LOINC2HPO_EPIC AND SCC

### 2022-04-27 code/r/01_MSDW_HPO_Phenotype_Analysis.Rmd
update

## GOAL
This is to LOINC2HPO analysis of both EPIC and SCC labs. 
```{r message=FALSE}
library(tidyverse)
library(glue)
library(readxl)
require(DBI)
require(RPostgres)
require(egg)
require(gridExtra)
dsca <- DBI::dbConnect(RPostgres::Postgres(), 
                       host=Sys.getenv('DSCA_PHI_HOST'), 
                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
                       port = Sys.getenv('DSCA_PHI_PORT'), 
                       user=Sys.getenv('DSCA_PHI_USERNAME'),
                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
```

@@

# recode all unknown gender as one category
patient_race_sex_distribution <- patient_race_sex_distribution %>%
  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
  mutate(gender = recode(gender, .missing = 'Unknown'))

# the following cleaning are no longer needed after updating the race mapping file
# recode "African American" as "Black or African American"
# patient_race_sex_distribution <- patient_race_sex_distribution %>%
#   mutate(race = recode(race, 'African American' = 'Black or African American'))
```

```{r}
countByGender <- patient_race_sex_distribution %>% 
  group_by(gender) %>% 
  summarise(n = sum(n))

plot_by_sex <- ggplot(countByGender) + 
  geom_bar(aes(x = 1, y = n / sum(n), fill = gender), stat = "identity") + 
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral", Indeterminant = 'Cyan', Unknown = 'Orange')) + 
  xlab("") + ylab("patient count") +
  coord_polar(theta = "y") + 
  theme_bw() + 
  theme(panel.grid = element_blank(), legend.position = c(0.25, 0.7), 
        axis.text = element_blank(), axis.ticks = element_blank(), 
        axis.title = element_text(size = 8), legend.text = element_text(size = 8), 
        legend.title = element_text(size = 10), 
        legend.background = element_rect(fill=alpha('white', 0.1)))

ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
plot_by_sex
```

@@

plot_by_race  <- ggplot(countByRace_Gender %>% 
         filter(gender %in% c('Male', 'Female')) %>% 
         mutate(race = as.factor(race), n = if_else(gender == "Male", n, -n))) + 
  geom_bar(aes(x = reorder(race, -n, FUN=sum), y = n, fill = gender, group=gender), stat="identity") + 
##  scale_y_continuous(breaks = seq(-40,40, 10), labels = abs(seq(-40,40, 10))) +
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
  scale_y_continuous(labels = abs) +
  xlab("") + ylab("patient count") +
  coord_flip() + 
  theme_bw() + 
  theme(panel.grid = element_blank(), legend.position = "right", 
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
        legend.background = element_rect(fill=alpha('white', 0.1)))

ggsave(filename = glue("{CACHE_DIR}/plotByRace.png"), plot = plot_by_race, width = 5, height = 2)
plot_by_race
```

@@

patient_birth_year_sex_distribution <- get_patient_birth_year_distribution(FALSE)

@@

plot_by_birth_year_sex <- patient_birth_year_sex_distribution %>% ungroup() %>% 
  filter(gender %in% c('Male', 'Female')) %>% 
  mutate(n = if_else(gender == 'Female', -n, n)) %>%
  mutate(birth_year_bucket = factor(birth_year_bucket, levels = rev(levels(birth_year_bucket)))) %>% 
  # change bucket name (1800,1931] to '<= 1931'
  # change bucket name (2011, 2021] to '(2011, 2020]' -- our database cutoff is 2020
  mutate(birth_year_bucket = recode(birth_year_bucket, '(1800,1931]' = '<=1931')) %>% 
  mutate(birth_year_bucket = recode(birth_year_bucket, '(2011,2021]' = '(2011,2020]')) %>%{
    ggplot(.) +
  geom_bar(aes(x = birth_year_bucket, y = n, fill = gender), stat = 'identity') + 
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
  scale_y_continuous(labels = abs) +
  xlab('') + ylab('patient count') +
  coord_flip() + 
      theme_bw() + 
  theme(panel.grid = element_blank(), legend.position = "right", 
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
        legend.background = element_rect(fill=alpha('white', 0.1)))
  }

ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.png"), plot = plot_by_birth_year_sex, width = 4, height = 3)
plot_by_birth_year_sex

```

>>>>

## GOAL
This is to LOINC2HPO analysis of both EPIC and SCC labs. 
```{r message=FALSE}
library(tidyverse)
library(glue)
library(readxl)
require(DBI)
require(RPostgres)
require(egg)
require(gridExtra)
source('functions.R')
dsca <- DBI::dbConnect(RPostgres::Postgres(), 
                       host=Sys.getenv('DSCA_PHI_HOST'), 
                       dbname = Sys.getenv('DSCA_PHI_DBNAME'), 
                       port = Sys.getenv('DSCA_PHI_PORT'), 
                       user=Sys.getenv('DSCA_PHI_USERNAME'),
                       password= Sys.getenv('DSCA_PHI_PASSWORD'))
```

@@

# recode all unknown gender as one category
patient_race_sex_distribution <- patient_race_sex_distribution %>%
  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
  mutate(gender = recode(gender, .missing = 'Unknown'))
```

```{r}
countByGender <- patient_race_sex_distribution %>% 
  group_by(gender) %>% 
  summarise(n = sum(n)) %>% 
  mutate(fraction = n / sum(n))

plot_by_sex <- ggplot(countByGender) + 
  geom_bar(aes(x = 1, y = n / sum(n), fill = gender), stat = "identity") + 
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral", Indeterminant = 'Cyan', Unknown = 'Orange')) + 
  xlab("") + ylab("patient count") +
  coord_polar(theta = "y") + 
  theme_az() + 
  theme(legend.position = c(0.25, 0.7), 
        axis.text = element_blank(), axis.ticks = element_blank(), 
        legend.background = element_rect(fill=alpha('white', 0.1)))

ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.pdf'), plot = plot_by_sex, width = 2.5, height = 2.5)
plot_by_sex
```

@@

plot_by_race  <- ggplot(countByRace_Gender %>% 
         filter(gender %in% c('Male', 'Female')) %>% 
         mutate(race = as.factor(race), n = if_else(gender == "Male", n, -n))) + 
  geom_bar(aes(x = reorder(race, -n, FUN=sum), y = n, fill = gender, group=gender), stat="identity") + 
##  scale_y_continuous(breaks = seq(-40,40, 10), labels = abs(seq(-40,40, 10))) +
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
  scale_y_continuous(labels = abs) +
  xlab("") + ylab("patient count") +
  coord_flip() + 
  theme_az() + 
  theme(legend.background = element_rect(fill=alpha('white', 0.1)))

ggsave(filename = glue("{CACHE_DIR}/plotByRace.png"), plot = plot_by_race, width = 5, height = 2)
ggsave(filename = glue("{CACHE_DIR}/plotByRace.pdf"), plot = plot_by_race, width = 5, height = 2)
plot_by_race
```
Fraction of each race:
```{r}
countByRace_Gender %>% 
  group_by(race) %>% 
  summarise(n = sum(n)) %>%
  mutate(fraction = n / sum(n)) %>% 
  arrange(desc(fraction))
```

@@

patient_birth_year_sex_distribution <- get_patient_birth_year_distribution(force_rerun = FORCE_RERUN)

@@

plot_by_birth_year_sex <- patient_birth_year_sex_distribution %>% ungroup() %>% 
  filter(gender %in% c('Male', 'Female')) %>% 
  mutate(n = if_else(gender == 'Female', -n, n)) %>%
  mutate(birth_year_bucket = factor(birth_year_bucket, levels = rev(levels(birth_year_bucket)))) %>% 
  # change bucket name (1800,1931] to '<= 1931'
  # change bucket name (2011, 2021] to '(2011, 2020]' -- our database cutoff is 2020
  mutate(birth_year_bucket = recode(birth_year_bucket, '(1800,1931]' = '<=1931')) %>% 
  mutate(birth_year_bucket = recode(birth_year_bucket, '(2011,2021]' = '(2011,2020]')) %>%{
    ggplot(.) +
  geom_bar(aes(x = birth_year_bucket, y = n, fill = gender), stat = 'identity') + 
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
  scale_y_continuous(labels = abs) +
  xlab('') + ylab('patient count') +
  coord_flip() + 
  theme_az() + 
  theme(legend.background = element_rect(fill=alpha('white', 0.1)))
  }

ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.png"), plot = plot_by_birth_year_sex, width = 4, height = 3)
ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.pdf"), plot = plot_by_birth_year_sex, width = 4, height = 3)
plot_by_birth_year_sex
```
